---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ebvnetcdf

<!-- badges: start -->
<!-- badges: end -->

This package can be used to easily access the data of the EBV NetCDFs which can be downloaded from the
[Geobon Portal](https://portal.geobon.org/). It also provides some basic visualization. Advanced users can build their own NetCDFs with the EBV standard.


## Installation

The ebvnetcdf packages is not yet released on CRAN. You can install it with: 

``` r
#visibility needs to be set to public first! Does not work right now. 
devtools::install_gitlab("lq39quba/ebvnetcdf", host='git.idiv.de') 
```

## Working with the package - a quick intro

### Take a very first look at the file

With these two functions you get a basic impression what data you can get from the EBV NetCDF.

```{r example}
library(ebvnetcdf)

file <- paste0(path.package("ebvnetcdf"),"/extdata/cSAR_idiv_v1.nc")

#take a look at some basic properties of that file
#prop.file <- ebv_properties(file)

#get all possible paths to the datacubes - dataframe including the paths and also descriptions of e.g. metric and or scenario - take a look!
#datacubes <- ebv_datacubepaths(file)

#get properties of one specific datacube - also holds the general file properties from above
#this time you get the warning that the value_range does not exists. So don't take the displayed value_range seriously.
#prop.dc <- ebv_properties(file, datacubes[1,1], verbose=T)

```

### Plot the data to get a better impression

```{r}
# Plot a map of the datacube that we just looked at - it has 12 timesteps, mabe look at two different ones?
# dc <- datacubes[1,1]
# ebv_plot_map(file, dc, timestep = 1)
# ebv_plot_map(file, dc, timestep = 6)

# what was the data about again? Check the properties!
# prop.dc@title
# And the datacube?
# prop.dc@entity_information@label

# It's nice to see the global distribution, but how it the change of that datacube (non forest birds) over time? Let's take a look at the average. The function returns the values, catch them!
# averages <- ebv_plot_indicator(file, dc)

# It would be cool to have that for other indicators as well? Well you have to wait for an update of the package. Or maybe implement it yourself using the functions coming up next?
```

### Read the data from the files to start working

### Take a peek on the creation of an EBV NetCDF



